Supplementary methods

### 2021-06-28 text/supplemental_methods.Rmd
modify supplemental methods section to reflect the newer model

We used Bayesian inference to explore the relative impact of cancer diagnosis on
survival at the time of diagnosis. Our model assumes that at birth each person
has a small initial probability per unit time to die, and that this probability
grows exponentially in time at some baseline rate that matches the observed CDC
data. The rate of growth in risk is assumed to jump to a higher level at
diagnosis, which we take as a proxy for onset. This represents an accelerated
aging type model. To fit the model, we constrain the baseline risks using the
available CDC life tables (ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also constrain the cancer specific multiplier
using the longevity data within the cancer tumor dataset. Our model estimates
the effect of cancer type and class-specific cancer effects independently,
resulting in a cancer and class specific estimate of the aging rate multiplier.

Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes
accuracy in the right tail of the distribution. We modeled baseline longevity
using a risk rate function:

@@

Time is split into two components. $A$ going from 0 up to the age of the patient
at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
is an indicator for whether diagnosis has occurred or not (before diagnosis
$D(t) = 0$, after $D(t) = 1$). The model assumes that the risk of death is
$R_{20}$ at age 20 years and increases exponentially with a constant rate $a$ up
to the age at diagnosis. Risk of death assumes a new rate $ka$ thereafter.
Though the model extends to earlier ages, relatively few patients were under age
20 (**SM2 Fig**).

@@

The $R(T)$ function has two important parameters, the $a$ value which represents
the background risk rates, and the $k$ value, which is a function of both cancer
tissue type, and the cancer cluster identity.

$$k = k_{tis} k_{cl}$$

The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
probability at 1 and a relatively narrow width, due to the fact that it is a
tissue specific multiplicative modifying factor. This allows the overall
magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior
is significantly less constrained so that there is a wider range of risk across
tissue types (**SM3 Fig**).

The $k$ in our model can be taken as, roughly, the aging rate relative to the
baseline rates. Thus, a value of $k=2$ means that a given cancer causes you to
experience the same risk in time $dt$ as a non-cancer patient would experience
at the same age, in time $2dt$.

@@

**Fig SM3. Priors for the tissue and class coefficients.** The tissue coefficient
is chosen to admit parameter values across a range from near 0 to 20+
in the high density region, reflecting the wide variety of severities
across tissue type. Within a given tissue, the class k parameter
multiplies the tissue k and is given a peak density at 1, and a
credible range from near 0 to 5, reflecting our expectation of a
moderate amount of variation across the classes.

>>>>

We used Bayesian inference to explore the relative impact of cancer diagnosis on
survival at the time of diagnosis. Our model assumes that at birth each person
has a small initial probability per unit time to die, and that this probability
grows exponentially in time at some baseline rate that matches the observed CDC
data. We assume that at diagnosis this information will allow us to treat
patients as if they were a different age, and with further time passing at a
different rate, a self-similarity based model. The effective age at diagnosis is
based on a baseline age for that tissue/class combination, plus a perturbation
for each year over 20 years old at diagnosis. To fit the model, we constrain the
baseline risks using the available CDC life tables
(ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also
constrain the cancer specific parameters using the longevity data within the
cancer tumor dataset. Our model estimates the effect of cancer type and
class-specific cancer effects independently, resulting in a cancer and class
specific estimate of the effective age.

Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes
accuracy in the right tail of the distribution. We modeled baseline longevity
using a risk rate function. The risk rate of death at a time $t$ days after the
$A$ birthday is given as:

@@

Time is split into two components. $A$ represents the "effective age" of the
patient at diagnosis, whereas $t$ starts at 0 upon diagnosis and represents the
days since diagnosis, a field recorded in the dataset. $D(t)$
is an indicator for whether diagnosis has occurred or not (before diagnosis
$D(t) = 0$, after $D(t) = 1$). The model assumes that the risk of death is
$R_{20}$ at age 20 years and increases exponentially with a constant rate $a$ up
to the age at diagnosis. Risk of death assumes a new rate $ka$ thereafter.

@@

However, to model the risk of cancer effectively we found that it was necessary
not to use the actual age $\A'$ but instead the "effective age" for that
tissue/class type. Our effective age is calculated as:

$$A = A_{0} A_{cl} + A_r (A'-20)$$ Where $A_0$ is the tissue specific effective age
parameter (expressed in years), $A_{cl}$ is a dimensionless multiplier for the
given class, and $A_r$ is the dimensionless age rate which determines how actual
age $A'$ affects effective age A using a linear perturbation.

The $R(T)$ function has two important parameters, the $a$ value which represents
the background risk rates, and the $k$ value, which is a function of cancer
tissue type.

$$k = k_{tis}$$

The product $A_0 A_{cl}$ is obviously symmetric between the two factors. To
disambiguate the meaning of the two, the prior for $A_0$ ranges from about 40 to
100 years, whereas the prior distribution of $A_{cl}$ has peak
probability at 1 and a relatively narrow width, due to the fact that it is a
tissue specific multiplicative modifying factor. To understand the roles of
these priors refer to (**SM3 Fig**).

@@

**Fig SM3. Priors for the tissue and class coefficients.** The tissue
coefficient is chosen to admit ages between 40 and 100 years old as a baseline
age. Within a given tissue, the class parameter multiplies the tissue value and
is given a peak density at 1, and a credible range from near 0.6 to 1.4,
reflecting our expectation of a moderate multiplicative perturbation of
variation across the classes.
